---
title: "Group A — Yogurt Study Data Dictionary"
subtitle: "Exploring the yogurt dataset: sample IDs, participant metadata, qPCR, and Luminex"
author: "Group A"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

## Overview

This document describes every CSV file in `data/group_a_yogurt/`.  
Each section shows what a **single row** represents, then explains every column.

The study is about the effect of **yogurt consumption** (vs. unchanged diet) on
the vaginal microbiome and immune environment, measured before and after a
course of antibiotics.

```{r setup, message=FALSE}
library(tidyverse)
```

---

## 00 — Sample IDs {#sec-sample-ids}

**File:** `00_sample_ids_yogurt.csv`

**One row represents:** one biological sample collected from one participant at
one time point. A single participant (`pid`) can appear twice (baseline and
after antibiotics).

```{r read-sample-ids}
sample_ids <- read_csv("../data/group_a_yogurt/00_sample_ids_yogurt.csv")
glimpse(sample_ids)
```

### Columns

| Column | Type | What it means |
|--------|------|---------------|
| `pid` | character | **Participant ID** — a unique code for each person in the study (e.g. `pid_25`). Links to the metadata file. |
| `time_point` | character | When the sample was taken: `"baseline"` (before antibiotics) or `"after_antibiotic"`. |
| `arm` | character | Which study group the participant was in: `"yogurt"` (ate yogurt during the study) or `"unchanged_diet"` (control group, no change to diet). |
| `sample_id` | character | **Lab sample ID** — a unique identifier for the physical sample that was sent to the lab. Starts with `UNCH` (unchanged diet) or `YOG` (yogurt). Used to link with qPCR and Luminex results. |

```{r peek-sample-ids}
# Quick peek: how many participants, samples per time point
sample_ids |> count(time_point, arm)
```

---

## 01 — Participant Metadata {#sec-metadata}

**File:** `01_participant_metadata_yogurt.csv`

**One row represents:** one participant enrolled in the study.

```{r read-metadata}
metadata <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
glimpse(metadata)
```

### Columns

| Column | Type | What it means |
|--------|------|---------------|
| `pid` | character | **Participant ID** — same code as in the sample IDs file. One row per person. |
| `arm` | character | Study arm: `"yogurt"` or `"unchanged_diet"` (control). |
| `days_since_last_sex` | integer | **Number of days** since the participant last had sex at the time of enrollment. |
| `birth_control` | character | Type of birth control used: `"no hormonal birth control"` or `"Depoprovera"` (an injectable hormonal contraceptive). |
| `age` | integer | Participant's age in years. |
| `education` | character | Highest education level: `"less than grade 9"`, `"grade 10-12, not matriculated"`, `"grade 10-12, matriculated"`, or `"post-secondary"`. |
| `sex` | character | Sex assigned at birth (all `"female"` in this study). |

```{r peek-metadata}
# Quick peek: age range and birth control distribution
metadata |>
  summarise(
    n = n(),
    min_age = min(age),
    max_age = max(age),
    mean_age = mean(age)
  )

metadata |> count(arm, birth_control)
```

---

## 02 — qPCR Results {#sec-qpcr}

**File:** `02_qpcr_results_yogurt.csv`

**One row represents:** the abundance (by qPCR) of three bacterial targets in
one vaginal sample.

qPCR (quantitative polymerase chain reaction) is a lab technique that measures
how much DNA of a specific bacterium is present. Higher numbers = more of that
bacterium.

```{r read-qpcr}
qpcr <- read_csv("../data/group_a_yogurt/02_qpcr_results_yogurt.csv")
glimpse(qpcr)
```

### Columns

| Column | Type | What it means |
|--------|------|---------------|
| `sample_id` | character | **Sample ID** — links to the sample IDs file (00). |
| `qpcr_bacteria` | double | **Total bacterial load** — total amount of bacterial DNA in the sample (higher = more bacteria overall). |
| `qpcr_crispatus` | double | Amount of *Lactobacillus crispatus* DNA. This is a "good" bacterium linked to a healthy vaginal microbiome. |
| `qpcr_iners` | double | Amount of *Lactobacillus iners* DNA. This species is less protective than *L. crispatus* and is often found in less stable microbiomes. |

```{r peek-qpcr}
# How many samples have detectable crispatus vs iners?
qpcr |>
  summarise(
    n_samples = n(),
    crispatus_detectable = sum(qpcr_crispatus > 0),
    iners_detectable = sum(qpcr_iners > 0)
  )
```

---

## 03 — Luminex Results {#sec-luminex}

**File:** `03_luminex_results_yogurt.csv`

**One row represents:** the concentration of one **cytokine** (immune protein)
measured in one vaginal sample. So each sample has multiple rows — one per
cytokine tested.

Luminex is a multiplex assay: it measures many different immune proteins in the
same liquid sample at once.

```{r read-luminex}
luminex <- read_csv("../data/group_a_yogurt/03_luminex_results_yogurt.csv")
glimpse(luminex)
```

### Columns

| Column | Type | What it means |
|--------|------|---------------|
| `sample_id` | character | **Sample ID** — links to the sample IDs file (00). |
| `cytokine` | character | Which immune protein was measured. Options: `IL-1a`, `IL-1b`, `IL-6`, `IL-8`, `IL-10`, `TNFa`, `IP-10`, `MIG`, `IFN-Y`, `MIP-3a`. These are signalling molecules the immune system uses to communicate. |
| `conc` | double | **Concentration** of that cytokine in the sample (units are pg/mL — picograms per millilitre). |
| `limits` | character | Whether the measured concentration was within the reliable detection range of the assay: `"within limits"` or `"out of range"` (too low or too high to be measured accurately). |

```{r peek-luminex}
# How many measurements per cytokine?
luminex |> count(cytokine)

# Which cytokines often go out of range?
luminex |>
  count(cytokine, limits) |>
  arrange(desc(n))
```

---

## Putting It All Together

These four files can be joined through the `sample_id` and `pid` columns:

1. **01 → 00:** join by `pid` to add participant background to each sample
2. **00 → 02:** join by `sample_id` to add bacterial abundance data
3. **00 → 03:** join by `sample_id` to add immune protein data

This lets you ask questions like:  
> "Does yogurt consumption change the levels of *L. crispatus* or inflammatory cytokines after antibiotics?"

```{r join-demo, message=FALSE}
# Example: combine everything into one wide dataset
samples_with_metadata <- sample_ids |>
  left_join(metadata, by = "pid")

# A glimpse at the joined data
samples_with_metadata |>
  select(pid, time_point, arm.x, age, birth_control) |>
  distinct() |>
  head(8)
```
