---
title: "Yogurt & the Vaginal Microbiome After Antibiotics — Group Report"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
execute:
  warning: false
  message: false
---

> A workshop project analysing how yogurt consumption affects the vaginal
> microbiome in women taking antibiotics.

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)
library(ggpubr)

# ── Read all 4 data files ──────────────────────────────────────────
samples  <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
metadata <- read_csv(
  here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"),
  col_types = cols(sex = col_character())
)
qpcr     <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))
luminex  <- read_csv(here("data/group_a_yogurt/03_luminex_results_yogurt.csv"))
```

---

## Q1 — Which variable defines the groups you want to compare?

```{r}
# The grouping variable is `arm`
metadata |>
  count(arm)
```

**Answer:** The variable **`arm`** defines the groups. It has two levels:

| Value | Meaning | N |
|---|---|---|
| `yogurt` | Ate yogurt daily after antibiotics | 26 |
| `unchanged_diet` | Kept usual diet (control) | 25 |

Every comparison in this report is **yogurt vs. unchanged_diet** at the
"after_antibiotic" time point (or baseline → after change).

---

## Q2 — Which demographic and baseline variables belong in your Table 1?

**Answer:** The variables that describe participants *before* the intervention
and could potentially differ between arms (even with randomisation):

| Variable | Type | Why it goes in Table 1 |
|---|---|---|
| `age` | Continuous | Could affect immune response and microbiome |
| `education` | Categorical | Proxy for socioeconomic factors |
| `birth_control` | Categorical | Hormonal BC affects the vaginal microbiome |
| `days_since_last_sex` | Continuous | Recent sex can temporarily alter the microbiome |
| ~~`sex`~~ | Constant | All female — skip it |
| ~~`pid`~~ | ID | Not a variable — skip it |

### Table 1

```{r}
# ── Table 1 ────────────────────────────────────────────────────────
metadata |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(
    test = list(age ~ "t.test")
  ) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

**Interpretation:**

- **Age, education, days_since_last_sex** — well balanced (p > 0.05). The
  randomisation did its job for these.
- **birth_control** — imbalanced (p = 0.017). The yogurt arm has twice as
  many Depoprovera users (65% vs 32%). This is worth noting because
  hormonal contraception can affect vaginal *Lactobacillus* levels. If
  we see a difference between arms in the outcomes, we may need to check
  whether birth control is driving it, not the yogurt.

---

## Q3 — Which outcome variables are most relevant to your question?

**Original question:** *How does yogurt affect the vaginal microbiome in
response to antibiotics?*

```{r}
# ── What qPCR outcomes are available ───────────────────────────────
qpcr |>
  select(-sample_id) |>
  names()
```

```{r}
# ── What cytokines are available ───────────────────────────────────
luminex |>
  distinct(cytokine) |>
  pull(cytokine)
```

**Answer:**

| Outcome | Data source | What it measures | Relevance |
|---|---|---|---|
| `qpcr_crispatus` | qPCR | *L. crispatus* — the "gold standard" protective species | **High** — should increase if yogurt helps |
| `qpcr_iners` | qPCR | *L. iners* — a less stable species, marker of dysbiosis | **High** — should decrease with recovery |
| `qpcr_bacteria` | qPCR | Total bacterial load | Medium — context |
| Cytokines (IL-1a, IL-6, IL-8, IL-1b, IP-10, MIG, etc.) | Luminex | Immune/inflammatory proteins in vaginal fluid | **High** — lower = less inflammation |

---

## Q4 — For each outcome, what kind of comparison makes sense?

```{r}
# ── Join qPCR with sample info to see what time points exist ───────
qpcr_joined <- qpcr |>
  left_join(samples, by = "sample_id")

qpcr_joined |>
  count(arm, time_point)
```

**Answer:** Two comparison types, each answering a different question:

### A) Independent groups at one time point (primary)

Compare `yogurt` vs `unchanged_diet` **at the after_antibiotic visit only**.

- **Question:** Did the yogurt group end up with a healthier microbiome?
- **Test:** Wilcoxon rank-sum test (two independent groups)
- **Applies to:** All outcomes (qPCR, cytokines)

### B) Paired pre-post within each arm

Compare each participant's *baseline* value to their *after_antibiotic* value,
separately for each arm.

- **Question:** Did each group change significantly from before to after?
- **Test:** Wilcoxon signed-rank test (paired)
- **Applies to:** qPCR outcomes only (we have both time points)

### Summary table

| Outcome | Comparison A (independent) | Comparison B (paired) |
|---|---|---|
| `qpcr_crispatus` | ✅ Primary: yogurt vs control at after_antibiotic | ✅ Did crispatus increase from baseline? |
| `qpcr_iners` | ✅ Secondary | ✅ Did iners decrease from baseline? |
| Cytokines | ✅ Primary: yogurt vs control at after_antibiotic | ❌ Only measured after_antibiotic |

---

## Q5 — Which `ggpubr` plot best shows that comparison?

### A) Independent groups — boxplot (crispatus) / Paired-change boxplot (iners)

Best for comparing yogurt vs control at after_antibiotic. For *L. iners* the
paired change (after_antibiotic − baseline) is shown as a boxplot with
individual points, following your preference.

```{r}
#| fig-width: 7
#| fig-height: 5
# ── Prepare qPCR data for plotting ─────────────────────────────
qpcr_after <- qpcr |>
  left_join(samples, by = "sample_id") |>
  filter(time_point == "after_antibiotic")

# ── L. crispatus: boxplot by arm ───────────────────────────────
ggboxplot(qpcr_after,
  x = "arm", y = "qpcr_crispatus",
  fill = "arm", palette = "jco",
  add = "jitter"
) +
  stat_compare_means(method = "wilcox.test", label.x = 1.3) +
  labs(
    y = expression(italic("L. crispatus") ~ "(copies / µL)"),
    title = expression(italic("L. crispatus") ~ "after antibiotics")
  ) +
  theme(legend.position = "none")
```

```{r}
#| fig-width: 7
#| fig-height: 5
# ── Compute paired change in L. iners ────────────────────────────
iners_change <- qpcr |>
  left_join(samples, by = "sample_id") |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  select(pid, arm, time_point, qpcr_iners) |>
  pivot_wider(names_from = time_point, values_from = qpcr_iners) |>
  mutate(change = after_antibiotic - baseline)

# ── Boxplot of the paired change, by arm ────────────────────────
ggboxplot(iners_change,
  x = "arm", y = "change",
  fill = "arm", palette = "jco",
  add = "jitter"
) +
  stat_compare_means(method = "wilcox.test", label.x = 1.3) +
  labs(
    y = expression(italic("L. iners") ~ "change (after − baseline, copies / µL)"),
    title = expression(italic("L. iners") ~ "— paired change after antibiotics")
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  theme(legend.position = "none")
```

### B) Paired pre-post — connected line plot

Best for showing change within each arm.

```{r}
#| fig-width: 9
#| fig-height: 5
# ── Prepare paired data ────────────────────────────────────────
qpcr_prepost <- qpcr |>
  left_join(samples, by = "sample_id") |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

# ── L. crispatus: paired plot ──────────────────────────────────
ggplot(qpcr_prepost, aes(time_point, qpcr_crispatus)) +
  geom_line(aes(group = pid), alpha = 0.3, color = "grey50") +
  geom_point(aes(color = arm), size = 2, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_color_manual(values = c("unchanged_diet" = "#E64B35", "yogurt" = "#4DABD7")) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "after_antibiotic")),
    label.y = max(qpcr_prepost$qpcr_crispatus, na.rm = TRUE) * 0.95
  ) +
  labs(
    x = "Time point",
    y = expression(italic("L. crispatus") ~ "(copies / µL)"),
    title = expression(italic("L. crispatus") ~ "— baseline → after antibiotics")
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```



### C) Total bacteria — boxplot by arm

Compares total bacterial load between the yogurt and unchanged-diet groups
after antibiotics. Each point is one participant.

```{r}
#| fig-width: 7
#| fig-height: 5
# ── Total bacteria: boxplot by arm ───────────────────────────
ggboxplot(qpcr_after,
  x = "arm", y = "qpcr_bacteria",
  fill = "arm", palette = "jco",
  add = "jitter"
) +
  stat_compare_means(method = "wilcox.test", label.x = 1.3) +
  labs(
    y = "Total bacteria (gene copies / µL)",
    title = "Total bacteria after antibiotics"
  ) +
  theme(legend.position = "none")
```

### D) Cytokine panel — faceted boxplots

Best for comparing multiple immune markers side by side.

```{r}
#| fig-width: 10
#| fig-height: 7
# ── Prepare cytokine data ──────────────────────────────────────
cytokines_keep <- c("IL-1a", "IL-6", "IL-8", "IL-1b", "MIG", "IP-10")

cytokine_df <- luminex |>
  filter(cytokine %in% cytokines_keep, limits == "within limits") |>
  left_join(samples, by = "sample_id") |>
  filter(time_point == "after_antibiotic")

# ── Faceted boxplot panel ──────────────────────────────────────
ggplot(cytokine_df, aes(arm, conc, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  facet_wrap(~ cytokine, scales = "free_y", nrow = 2) +
  scale_fill_manual(values = c("unchanged_diet" = "#E64B35", "yogurt" = "#4DABD7")) +
  stat_compare_means(method = "wilcox.test", label.x = 1.3, size = 3) +
  labs(
    x = "Arm",
    y = "Concentration (pg/mL)",
    title = "Vaginal cytokines after antibiotics"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```

### E) Bacteria boxplots — both time points, by arm

For each bacterial target, a boxplot showing baseline on the left and
after_antibiotic on the right, with the two arms side by side at each time
point. P-values compare unchanged_diet vs yogurt within each time point.

```{r}
#| fig-width: 9
#| fig-height: 14
#| message: false
# ── Prepare data ───────────────────────────────────────────────
qpcr_data <- qpcr |>
  left_join(samples, by = "sample_id") |>
  filter(time_point %in% c("baseline", "after_antibiotic")) |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

# ── Helper: one boxplot for a given bacteria column ────────────
make_bp <- function(data, y_col, y_label, title_text) {
  ggplot(data, aes(time_point, {{ y_col }}, fill = arm)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.7) +
    geom_jitter(aes(color = arm), width = 0.15, alpha = 0.4, size = 1.2) +
    scale_fill_manual(values = c("unchanged_diet" = "#E64B35", "yogurt" = "#4DABD7")) +
    scale_color_manual(values = c("unchanged_diet" = "#E64B35", "yogurt" = "#4DABD7")) +
    stat_compare_means(aes(group = arm), label = "p.format", method = "wilcox.test") +
    labs(x = "Time point", y = y_label, title = title_text) +
    theme_minimal() +
    theme(legend.position = "bottom")
}

# ── Three boxplots stacked vertically ──────────────────────────
library(patchwork)

p_bact <- make_bp(qpcr_data, qpcr_bacteria,
  "Gene copies / \u00B5L", "Total bacteria")
p_cris <- make_bp(qpcr_data, qpcr_crispatus,
  expression(italic("L. crispatus") ~ "(copies / \u00B5L)"),
  expression(italic("L. crispatus")))
p_iners <- make_bp(qpcr_data, qpcr_iners,
  expression(italic("L. iners") ~ "(copies / \u00B5L)"),
  expression(italic("L. iners")))

p_bact / p_cris / p_iners
```

---

## Q6 — What does a p-value mean in the context of your comparison?

### What it *does* tell you

> If yogurt truly had **no effect** (the null hypothesis is true), how likely
> would we be to see a difference this big (or bigger) just by random chance?

**Example:** If the *crispatus* boxplot shows p = 0.03, it means:
> "There's only a 3% chance of seeing this much difference between the
> yogurt and control groups if yogurt actually does nothing."

| If p is ... | We say ... |
|---|---|
| p < 0.05 | "Statistically significant" — unlikely to be chance |
| p ≥ 0.05 | "Not statistically significant" — could be chance |

### What a p-value does **NOT** tell you

| ❌ It does NOT mean ... | Why ... |
|---|---|
| "The effect is large" | A tiny effect can be significant with enough people |
| "The effect is important" | Statistical significance ≠ clinical/practical importance |
| "Yogurt definitely works" | p-values can't prove causation |
| "There's a 95% chance yogurt works" | That's a different quantity (Bayesian, not frequentist) |
| "The result will replicate" | One study, one p-value — replication is separate |
| "The study was well-designed" | Confounding, bias, small samples all affect what p means |

### In your study specifically

Even if the *crispatus* p-value is > 0.05 (not significant), it doesn't mean
"yogurt didn't help." With only 51 participants, the study may simply be
**underpowered** — too small to detect a real but modest effect.

Conversely, if the *birth control imbalance* (p = 0.017 in Table 1) is real,
it could **confound** the results. A significant p-value for *crispatus*
might partly reflect the hormonal BC difference, not the yogurt.

---

## Q7 — What surprised you in the results?

```{r}
# ── Quick summary stats to look for surprises ──────────────────
qpcr_after |>
  group_by(arm) |>
  summarise(
    n = n(),
    crispatus_median = median(qpcr_crispatus, na.rm = TRUE),
    crispatus_zero_pct = mean(qpcr_crispatus == 0, na.rm = TRUE) * 100,
    iners_median = median(qpcr_iners, na.rm = TRUE)
  )
```

```{r}
# ── How many participants have zero crispatus? ─────────────────
qpcr_after |>
  count(arm, zero_crispatus = qpcr_crispatus == 0)
```

**Check these things and note what surprises you below:**

1. **Zero inflation** — How many women have *zero* detectable *L. crispatus*
   after antibiotics? If it's most of them, that's a striking finding.

2. **Outliers** — Are there one or two participants with extremely high values
   that might be driving a significant p-value all by themselves?

3. **The birth control imbalance** — Could the Depoprovera imbalance in Table 1
   (p = 0.017) explain any arm differences you see in the outcomes?

4. **Cytokine surprises** — Are any cytokines *higher* in the yogurt group
   than the control (counterintuitive)? Which ones?

5. **Median vs mean** — If the median and mean are very different, the data
   are skewed. The Wilcoxon test is better for skewed data than a t-test.

```{r}
# ── Write your surprises here (plain text) ─────────────────────
```

> **Your observations:**
>
> 1.
> 2.
> 3.

---

## Summary

| Question | Answer |
|---|---|
| **Grouping variable** | `arm` (yogurt vs unchanged_diet) |
| **Table 1 variables** | age, education, birth_control, days_since_last_sex |
| **Key outcomes** | `qpcr_crispatus`, `qpcr_iners`, cytokines |
| **Primary comparison** | Independent groups at after_antibiotic (Wilcoxon rank-sum) |
| **Secondary comparison** | Paired pre-post within each arm (Wilcoxon signed-rank) |
| **Best plots** | Boxplot + jitter (independent); connected line plot (paired); faceted boxplot panel (cytokines) |
| **P-value limits** | Tells you about chance, not about size, importance, or causation |
 ```r                                                                                          
   library(tidyverse)                                                                          
   library(ggpubr)                                                                             
                                                                                               
   # Read data                                                                                 
   samples  <- read_csv("data/group_a_yogurt/00_sample_ids_yogurt.csv")                        
   qpcr     <- read_csv("data/group_a_yogurt/02_qpcr_results_yogurt.csv")                      
                                                                                               
   # Compute paired change (after_antibiotic - baseline)                                       
   iners_change <- qpcr |>                                                                     
     left_join(samples, by = "sample_id") |>                                                   
     filter(time_point %in% c("baseline", "after_antibiotic")) |>                              
     select(pid, arm, time_point, qpcr_iners) |>                                               
     pivot_wider(names_from = time_point, values_from = qpcr_iners) |>                         
     mutate(change = after_antibiotic - baseline)                                              
                                                                                               
   # Boxplot of the paired change                                                              
   ggboxplot(iners_change,                                                                     
     x = "arm", y = "change",                                                                  
     fill = "arm", palette = "jco",                                                            
     add = "jitter"                                                                            
   ) +                                                                                         
     stat_compare_means(method = "wilcox.test", label.x = 1.3) +                               
     labs(                                                                                     
       y = expression(italic("L. iners") ~ "change (after - baseline, copies / µL)"),          
       title = expression(italic("L. iners") ~ "- paired change after antibiotics")            
     ) +                                                                                       
     geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +                       
     theme(legend.position = "none")                                                           
 ``` 
 