---
title: "Module 6 — Table 1 (Yogurt Group)"
format: html
editor: source
execute:
  warning: false
  message: false
---

> *Your mode for this exercise.* Group A: **agentic** — describe each step to Pi
> and let it write the code. Group B: **manual** — write each step yourself.

```{r setup}
library(tidyverse)
library(here)
library(gtsummary)

metadata <- read_csv(
  here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"),
  col_types = cols(sex = col_character())
)
```

---

## The data

```{r}
metadata
```

Your columns:

| Column | Type | What it is |
|---|---|---|
| `pid` | ID | Participant identifier (not a variable) |
| `arm` | **Group** | `yogurt` or `unchanged_diet` |
| `age` | Continuous | Years |
| `education` | Categorical | 4 levels: less than grade 9 / grade 10-12... |
| `birth_control` | Categorical | Depoprovera or no hormonal birth control |
| `days_since_last_sex` | Continuous | Days |
| `sex` | Constant | All female (skip this one) |

---

## Step 1 — Default `tbl_summary`

Start here. We drop `pid` and `sex` (sex is constant, pid is an ID):

```{r}
# step 1: tbl_summary defaults
metadata |>
  select(-pid, -sex) |>
  tbl_summary(by = arm)
```

**What to notice:** gtsummary automatically picks the right summary for each
variable type — median (IQR) for continuous, n (%) for categorical. It also
automatically picks the right test for `add_p()` (more on that below).

---

## Step 2 — Add p-values

Add `|> add_p()` to the chain:

```{r}
# step 2: add p-values
metadata |>
  select(-pid, -sex) |>
  tbl_summary(by = arm) |>
  add_p()
```

**What the p-values mean here:** These tests check whether the *baseline*
characteristics differ between the two arms. In a randomised study, we expect
most to be non-significant (p > 0.05), because randomisation should balance the
groups. A significant p-value would be a red flag — it might mean randomisation
didn't work perfectly, or the difference is just bad luck (that's why we
usually don't "adjust" for these, we just report them).

**Which test does `add_p()` pick?** It depends on the variable type:
- *Continuous* (`age`, `days_since_last_sex`): **Wilcoxon rank-sum test**
  (Mann-Whitney U) — a non-parametric test that doesn't assume normality.
- *Categorical* (`education`, `birth_control`, `arm`): **Fisher's exact test**
  (when any expected count < 5) or **chi-squared test**.

---

## Step 3 — Override `age` to mean (SD) + t-test

Sometimes people prefer mean ± SD for age (it's conventional in many fields).
You can override both the summary statistic and the test:

```{r}
# step 3: override age to mean (SD) + t-test
metadata |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(
    test = list(age ~ "t.test")
  )
```

`statistic` controls how each variable is displayed. `test` controls which
hypothesis test `add_p()` runs. You can mix and match — maybe you want
median (IQR) for `days_since_last_sex` but mean (SD) for `age`.

---

## Step 4 — Polish

Final touches: add an overall column, bold the labels, clean up the header:

```{r}
# step 4: polished table
metadata |>
  select(-pid, -sex) |>
  tbl_summary(
    by = arm,
    statistic = list(age ~ "{mean} ({sd})")
  ) |>
  add_p(
    test = list(age ~ "t.test")
  ) |>
  add_overall() |>
  bold_labels() |>
  modify_header(label ~ "**Characteristic**")
```

---

## Step 5 — Your interpretation

Now look at the p-values in your final table and answer:

1. **Age** — Are the yogurt and unchanged-diet groups similar in age? What's the p-value?
2. **Education** — Does one group have more participants with post-secondary education?
3. **Birth control** — Is Depoprovera use balanced between arms?
4. **Days since last sex** — Any imbalance here? (This one is sometimes not perfectly balanced even in randomised trials.)

If all p-values are > 0.05, the randomisation did its job — the groups look
comparable at baseline. Any differences you see *after* the intervention can
more confidently be attributed to the yogurt, not to pre-existing differences.

---

## Reach (optional)

Try building Table 1 with `by = education` instead of `by = arm`. What question
does *that* table answer?

```{r}
# reach: by = education
```