---
title: "Group A Yogurt Study — Paired Bacteria Analysis"
subtitle: "How bacteria levels changed from baseline to after antibiotics, by arm"
format: html
editor: source
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggpubr)
library(rstatix)
library(gtsummary)
library(here)
```

---

## Study Notes (read this first!)

### Aim of the study

The main goal of this study was to find out whether **eating yoghurt**
helps protect the vaginal microbes (bacteria) when a person takes
antibiotics. Antibiotics can kill both the bad bacteria causing an
infection AND the good bacteria that keep the vagina healthy, which can
lead to problems like yeast infections or bacterial vaginosis. The idea
is that yoghurt (which contains good bacteria like *Lactobacillus*)
might help restore or protect the vaginal microbiome.

### Objectives

1. Measure the levels of key vaginal bacteria at **baseline** (before
   any antibiotics) and **after taking antibiotics**
2. Compare what happens in two groups of participants:
   - **Unchanged diet** — people who went about their normal diet
   - **Yoghurt** — people who ate yoghurt during the study
3. See if the yoghurt group had better preservation of "good" bacteria
   (*L. crispatus*) compared to the unchanged diet group
4. Use statistical tests (p-values) to decide if any differences we see
   are real or just by chance

### How the study was conducted

Here's a simple walkthrough of the study design:

```
Step 1: Recruit participants (total N = 51 women)
                |
Step 2: Divide into two arms
        ├── Unchanged diet (~25 women) — eat as usual
        └── Yoghurt (~26 women) — eat yoghurt daily
                |
Step 3: Take baseline sample (swab) → measure bacteria
                |
Step 4: Everyone takes antibiotics (for an infection)
                |
Step 5: Take after-antibiotic sample → measure bacteria again
                |
Step 6: Compare: Did bacteria change? Did yoghurt help?
```

**Key idea — "Paired" design:** Each woman is measured twice (before
and after antibiotics). This is powerful because each person acts as
her **own control**. We're not comparing different people — we're
comparing each person to herself before and after. This removes a lot
of natural differences between people (everyone's microbiome is a
bit different).

### What we measured (the bacteria)

The researchers used a technique called **qPCR** (quantitative PCR) to
count how many copies of bacterial DNA were in each sample. Think of it
like a microscope that can count individual bacteria. They looked at:

| Bacteria column | What it means | Why it matters |
|----------------|---------------|----------------|
| `qpcr_bacteria` | **Total bacteria** — all bacteria combined | Gives us a sense of the overall bacterial community |
| `qpcr_crispatus` | ***Lactobacillus crispatus*** — a "good" bacteria | This species produces hydrogen peroxide which kills bad bacteria. High levels = healthy vaginal environment |
| `qpcr_iners` | ***Lactobacillus iners*** — another *Lactobacillus* species | Less protective than *crispatus*. Can switch between helpful and harmful depending on conditions |

### How we made the graphs (explained simply)

**Step 1 — Gather the data:** We read three CSV files — one with
the participant IDs and which group they were in, one with the
time point (baseline vs after antibiotic), and one with the qPCR
bacteria counts. We merged them together so every row has:
*who* the person is + *when* the sample was taken + *which group*
they belong to + *how much* bacteria was found.

**Step 2 — Organise the data:** We reshaped the data so each bacteria
(Total, *L. crispatus*, *L. iners*) becomes its own row. This makes
it easier to create one graph per bacteria.

**Step 3 — Create the plots:** For each bacteria, we drew:
- **Grey lines** connecting each person's two measurements (baseline
  → after antibiotic). Each line is one woman. If the line goes down,
  her bacteria decreased. If it goes up, it increased.
- **Box plots** showing the overall pattern for the group. The box
  shows the middle 50% of values, the line inside is the median
  (middle value), and the "whiskers" show the range.
- **Dots** for each individual measurement, coloured by arm.
- **P-values** calculated using a paired Wilcoxon test. This is a
  statistical test that asks: "Is the change from before to after
  antibiotics big enough that we can be confident it's not just
  random noise?"

**Step 4 — Build the tables:** For each bacteria and each arm, we
counted how many participants had detectable levels (n, %), calculated
the median (middle value) and IQR (spread), and added the p-value.

### Understanding p-values (simple explanation)

A p-value tells you the probability that what you're seeing happened
by random chance alone. 

- **p < 0.05** — less than 5% chance it's random. We call this
  "statistically significant". The difference is probably real.
- **p < 0.01** — less than 1% chance. Strong evidence of a real
  difference.
- **p < 0.001** — less than 0.1% chance. Very strong evidence.
- **p > 0.05** — not statistically significant. The difference could
  easily be due to random chance. We can't confidently say there's
  a real difference.

Note: A high p-value doesn't mean "no difference". It means "we don't
have enough evidence to say there IS a difference". With only ~25
people per group, sometimes real differences are missed because the
sample is too small.

### Advantages of this study design

| Advantage | Why it's good |
|-----------|---------------|
| **Paired design** | Each person is their own control. This removes the natural variation between different people, making it easier to spot real changes |
| **Two time points** | We see the *change* over time, not just a snapshot. This tells us about the effect of antibiotics + yoghurt |
| **qPCR measurement** | qPCR is very precise — it can detect tiny amounts of bacterial DNA, giving accurate counts |
| **Wilcoxon test** | This test doesn't assume the data follows a bell curve (normal distribution). Our data has many zeros and is very skewed, so this is the right choice |
| **Detection rates** | We show both how many people have the bacteria AND how much. This gives a complete picture |

### Disadvantages and limitations

| Limitation | Why it matters |
|------------|----------------|
| **Small sample size** | About 25 people per arm. With small groups, it's harder to detect real differences (low "statistical power") |
| **No randomisation info** | We don't know how people were assigned to yoghurt vs unchanged diet. If it wasn't random, the groups might be different in ways that affect the results |
| **Short follow-up** | Only one time point after antibiotics. We don't know if changes lasted or if people recovered |
| **No placebo** | People knew whether they were eating yoghurt or not. This could affect behaviour (placebo effect) |
| **Many zeros** | *L. crispatus* especially has many zero values (not detected). This makes statistical analysis harder and means median values are often 0 |
| **Single measurement method** | Only qPCR was used. Other methods (like 16S sequencing) might give a broader picture of all bacteria present |
| **Confounding factors** | Things like age, birth control, sexual activity, and diet outside yoghurt could influence the results. We haven't adjusted for these in these simple plots |

### Summary and key takeaways

Based on the graphs and tables in this document:

1. **Total bacteria** — After antibiotics, total bacteria levels tend
   to decrease in both groups. The p-values will tell us if the
   decrease is significant and whether the yoghurt group fared better.

2. ***L. crispatus*** — This is the key "good" bacteria we care about.
   Antibiotics often wipe it out. If yoghurt helps, we might see
   *crispatus* levels dropping less (or recovering more) in the
   yoghurt group. However, many participants start with zero, which
   makes it hard to show protection.

3. ***L. iners*** — This species is more mixed in its effects. It
   might increase after antibiotics as it fills the space left by
   other bacteria. The yoghurt group might show different patterns.

4. **Overall** — The paired plots (grey lines connecting before →
   after) are the most informative: they show individual trajectories.
   Look for patterns where most lines go in one direction (all down
   or all up) — that's a strong signal. If lines go every which way,
   there's no clear effect.

**What to look for in each graph:**
- Do most grey lines go down? → Bacteria decreased after antibiotics
- Is the decrease smaller in the yoghurt group? → Yoghurt might help
- Is the p-value < 0.05? → The change is statistically significant
- Are the detection rates similar at both time points? → Bacteria were
  preserved

### Why this matters

Vaginal health is important for preventing infections, especially in
women. If something as simple as eating yoghurt can protect the vaginal
microbiome during antibiotic treatment, it could be an easy,
low-cost recommendation for doctors to give their patients. This study
is a step toward understanding that question.

---

## About this file

This document shows how bacteria changed from **baseline** to **after
antibiotic** for each participant, split by diet arm (**unchanged
diet** vs **yogurt**). Each section has:

1. A **table** with participant numbers, percentages, bacteria
   summaries, and p-values
2. A **paired box plot** showing each person's change over time

### What the bacteria columns mean

| Column | What it measures |
|--------|-------------------|
| `qpcr_bacteria` | **Total bacteria** — all bacteria together |
| `qpcr_crispatus` | ***Lactobacillus crispatus*** — a "good" bacteria linked to vaginal health |
| `qpcr_iners` | ***Lactobacillus iners*** — another common species, less protective |

```{r load-data}
# Read the three data files
samples <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
meta    <- read_csv(here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
qpcr    <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))

glimpse(qpcr)
```

```{r merge-and-tidy}
# Merge qPCR values with time_point and arm info
df <- samples %>%
  left_join(qpcr, by = "sample_id")

# Clean up labels for nice-looking plots
df <- df %>%
  mutate(
    arm = recode(arm,
                 "unchanged_diet" = "Unchanged diet",
                 "yogurt"         = "Yogurt"),
    time_point = recode(time_point,
                        "baseline"         = "Baseline",
                        "after_antibiotic" = "After antibiotic"),
    # Put time points in the right order (baseline first)
    time_point = fct_relevel(time_point, "Baseline", "After antibiotic")
  )

# Pivot so each bacteria gets its own row
df_long <- df %>%
  pivot_longer(
    cols      = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to  = "bacteria",
    values_to = "value"
  ) %>%
  mutate(
    bacteria = recode(bacteria,
                      "qpcr_bacteria"  = "Total bacteria",
                      "qpcr_crispatus" = "L. crispatus",
                      "qpcr_iners"     = "L. iners"),
    # Keep the right order
    bacteria = fct_relevel(bacteria, "Total bacteria", "L. crispatus", "L. iners"),
    # Flag whether bacteria was detected (value > 0)
    detected = if_else(value > 0, "Detected", "Not detected")
  )

# Quick check: does each pid appear at both time points?
df_long %>%
  group_by(bacteria, pid) %>%
  summarise(n_timepoints = n_distinct(time_point), .groups = "drop") %>%
  count(n_timepoints)

head(df_long)
```

```{r helper-function}
# Helper function: compute paired Wilcoxon p-values per arm
compute_paired_pvals <- function(data) {
  paired_data <- data %>%
    select(pid, arm, bacteria, time_point, value) %>%
    pivot_wider(names_from = time_point, values_from = value)

  paired_data %>%
    group_by(arm) %>%
    summarise(
      p = wilcox.test(Baseline, `After antibiotic`, paired = TRUE, exact = FALSE)$p.value,
      .groups = "drop"
    ) %>%
    mutate(
      label = paste0("Paired Wilcoxon p = ", formatC(p, format = "e", digits = 2)),
      x_pos = 2.3
    )
}
```

---

## Overall participant summary

This first table shows how many participants are in each arm, plus the
total number of participants in the study.

```{r participant-counts}
# Number of unique participants per arm
participant_counts <- df_long %>%
  distinct(pid, arm, bacteria) %>%
  count(arm, bacteria) %>%
  mutate(pct = round(n / sum(n) * 100, 1),
         label = paste0(n, " (", pct, "%)"))

# Overall counts
overall_counts <- df_long %>%
  distinct(pid) %>%
  nrow()

overall_arm_counts <- df_long %>%
  distinct(pid, arm) %>%
  count(arm) %>%
  mutate(pct = round(n / overall_counts * 100, 1),
         label = paste0(n, " (", pct, "%)"))

# Display
knitr::kable(
  overall_arm_counts %>% select(arm, label),
  col.names = c("Arm", "Participants, n (%)"),
  caption = paste0("Table 0: Participants per arm (total N = ", overall_counts, ")")
)
```

---

## Total bacteria — table + paired plot

### Table: Total bacteria by arm

```{r table-total-bacteria}
# Widen data so each row is one participant with baseline + after_antibiotic
total_wide <- df_long %>%
  filter(bacteria == "Total bacteria") %>%
  select(pid, arm, time_point, value, detected) %>%
  pivot_wider(
    names_from  = time_point,
    values_from = c(value, detected),
    names_glue  = "{time_point}_{.value}"
  )

# Participants with detectable bacteria at each time point
total_detect <- total_wide %>%
  group_by(arm) %>%
  summarise(
    n_total       = n(),
    baseline_detected  = sum(Baseline_detected == "Detected"),
    baseline_pct       = round(baseline_detected / n_total * 100, 0),
    after_detected     = sum(`After antibiotic_detected` == "Detected"),
    after_pct          = round(after_detected / n_total * 100, 0),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_label = paste0(baseline_detected, "/", n_total, " (", baseline_pct, "%)"),
    after_label    = paste0(after_detected, "/", n_total, " (", after_pct, "%)")
  )

# Median (IQR) at each time point
total_summary <- total_wide %>%
  group_by(arm) %>%
  summarise(
    baseline_median_iqr = paste0(
      format(median(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    after_median_iqr = paste0(
      format(median(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    .groups = "drop"
  )

# Paired p-value
total_pval_arm <- total_wide %>%
  group_by(arm) %>%
  summarise(
    p = wilcox.test(Baseline_value, `After antibiotic_value`, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_label = formatC(p, format = "e", digits = 2))

# Combine into one table
total_table <- total_detect %>%
  left_join(total_summary, by = "arm") %>%
  left_join(total_pval_arm, by = "arm") %>%
  select(
    arm,
    n_total,
    baseline_label,
    baseline_median_iqr,
    after_label,
    after_median_iqr,
    p_label
  ) %>%
  rename(
    "Arm"                  = arm,
    "N"                    = n_total,
    "Baseline detected, n (%)" = baseline_label,
    "Baseline median (IQR)"    = baseline_median_iqr,
    "After antibiotic detected, n (%)" = after_label,
    "After antibiotic median (IQR)"    = after_median_iqr,
    "Paired p-value"             = p_label
  )

knitr::kable(total_table, caption = "Table 1: Total bacteria — detection rates and levels by arm")
```

### Plot: Total bacteria paired change

```{r plot-total-bacteria}
#| fig-width: 9
#| fig-height: 5

total_df <- df_long %>% filter(bacteria == "Total bacteria")
total_pvals <- compute_paired_pvals(total_df)

total_ymax <- total_df %>%
  group_by(arm) %>%
  summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop")

total_pvals <- total_pvals %>% left_join(total_ymax, by = "arm")

ggplot(total_df, aes(x = time_point, y = value)) +
  geom_line(aes(group = pid), color = "grey60", alpha = 0.5, linewidth = 0.7) +
  geom_boxplot(aes(fill = arm), outlier.shape = NA, alpha = 0.4, width = 0.4) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  scale_color_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = total_pvals,
    aes(x = x_pos, y = max_val, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Total bacteria",
    subtitle = "Paired change from baseline to after antibiotics",
    x        = NULL,
    y        = "Bacterial load (copies/mL)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "none"
  )
```

**Reading this:** Each grey line is one participant. Sloping **down**
means their bacteria went down after antibiotics. The p-value tests if
the overall change is different from zero.

---

## *Lactobacillus crispatus* — table + paired plot

### Table: *L. crispatus* by arm

```{r table-crispatus}
crispatus_wide <- df_long %>%
  filter(bacteria == "L. crispatus") %>%
  select(pid, arm, time_point, value, detected) %>%
  pivot_wider(
    names_from  = time_point,
    values_from = c(value, detected),
    names_glue  = "{time_point}_{.value}"
  )

crispatus_detect <- crispatus_wide %>%
  group_by(arm) %>%
  summarise(
    n_total            = n(),
    baseline_detected  = sum(Baseline_detected == "Detected"),
    baseline_pct       = round(baseline_detected / n_total * 100, 0),
    after_detected     = sum(`After antibiotic_detected` == "Detected"),
    after_pct          = round(after_detected / n_total * 100, 0),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_label = paste0(baseline_detected, "/", n_total, " (", baseline_pct, "%)"),
    after_label    = paste0(after_detected, "/", n_total, " (", after_pct, "%)")
  )

crispatus_summary <- crispatus_wide %>%
  group_by(arm) %>%
  summarise(
    baseline_median_iqr = paste0(
      format(median(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    after_median_iqr = paste0(
      format(median(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    .groups = "drop"
  )

crispatus_pval_arm <- crispatus_wide %>%
  group_by(arm) %>%
  summarise(
    p = wilcox.test(Baseline_value, `After antibiotic_value`, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_label = formatC(p, format = "e", digits = 2))

crispatus_table <- crispatus_detect %>%
  left_join(crispatus_summary, by = "arm") %>%
  left_join(crispatus_pval_arm, by = "arm") %>%
  select(arm, n_total, baseline_label, baseline_median_iqr,
         after_label, after_median_iqr, p_label) %>%
  rename(
    "Arm"                               = arm,
    "N"                                 = n_total,
    "Baseline detected, n (%)"          = baseline_label,
    "Baseline median (IQR)"             = baseline_median_iqr,
    "After antibiotic detected, n (%)"  = after_label,
    "After antibiotic median (IQR)"     = after_median_iqr,
    "Paired p-value"                    = p_label
  )

knitr::kable(crispatus_table,
             caption = "Table 2: L. crispatus — detection rates and levels by arm")
```

### Plot: *L. crispatus* paired change

```{r plot-crispatus}
#| fig-width: 9
#| fig-height: 5

crispatus_df <- df_long %>% filter(bacteria == "L. crispatus")
crispatus_pvals <- compute_paired_pvals(crispatus_df)

crispatus_ymax <- crispatus_df %>%
  group_by(arm) %>%
  summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop")

crispatus_pvals <- crispatus_pvals %>% left_join(crispatus_ymax, by = "arm")

ggplot(crispatus_df, aes(x = time_point, y = value)) +
  geom_line(aes(group = pid), color = "grey60", alpha = 0.5, linewidth = 0.7) +
  geom_boxplot(aes(fill = arm), outlier.shape = NA, alpha = 0.4, width = 0.4) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  scale_color_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = crispatus_pvals,
    aes(x = x_pos, y = max_val, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Lactobacillus crispatus",
    subtitle = "Paired change from baseline to after antibiotics",
    x        = NULL,
    y        = "Bacterial load (copies/mL)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "none"
  )
```

---

## *Lactobacillus iners* — table + paired plot

### Table: *L. iners* by arm

```{r table-iners}
iners_wide <- df_long %>%
  filter(bacteria == "L. iners") %>%
  select(pid, arm, time_point, value, detected) %>%
  pivot_wider(
    names_from  = time_point,
    values_from = c(value, detected),
    names_glue  = "{time_point}_{.value}"
  )

iners_detect <- iners_wide %>%
  group_by(arm) %>%
  summarise(
    n_total            = n(),
    baseline_detected  = sum(Baseline_detected == "Detected"),
    baseline_pct       = round(baseline_detected / n_total * 100, 0),
    after_detected     = sum(`After antibiotic_detected` == "Detected"),
    after_pct          = round(after_detected / n_total * 100, 0),
    .groups = "drop"
  ) %>%
  mutate(
    baseline_label = paste0(baseline_detected, "/", n_total, " (", baseline_pct, "%)"),
    after_label    = paste0(after_detected, "/", n_total, " (", after_pct, "%)")
  )

iners_summary <- iners_wide %>%
  group_by(arm) %>%
  summarise(
    baseline_median_iqr = paste0(
      format(median(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    after_median_iqr = paste0(
      format(median(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      " (",
      format(IQR(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
      ")"
    ),
    .groups = "drop"
  )

iners_pval_arm <- iners_wide %>%
  group_by(arm) %>%
  summarise(
    p = wilcox.test(Baseline_value, `After antibiotic_value`, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(p_label = formatC(p, format = "e", digits = 2))

iners_table <- iners_detect %>%
  left_join(iners_summary, by = "arm") %>%
  left_join(iners_pval_arm, by = "arm") %>%
  select(arm, n_total, baseline_label, baseline_median_iqr,
         after_label, after_median_iqr, p_label) %>%
  rename(
    "Arm"                               = arm,
    "N"                                 = n_total,
    "Baseline detected, n (%)"          = baseline_label,
    "Baseline median (IQR)"             = baseline_median_iqr,
    "After antibiotic detected, n (%)"  = after_label,
    "After antibiotic median (IQR)"     = after_median_iqr,
    "Paired p-value"                    = p_label
  )

knitr::kable(iners_table,
             caption = "Table 3: L. iners — detection rates and levels by arm")
```

### Plot: *L. iners* paired change

```{r plot-iners}
#| fig-width: 9
#| fig-height: 5

iners_df <- df_long %>% filter(bacteria == "L. iners")
iners_pvals <- compute_paired_pvals(iners_df)

iners_ymax <- iners_df %>%
  group_by(arm) %>%
  summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop")

iners_pvals <- iners_pvals %>% left_join(iners_ymax, by = "arm")

ggplot(iners_df, aes(x = time_point, y = value)) +
  geom_line(aes(group = pid), color = "grey60", alpha = 0.5, linewidth = 0.7) +
  geom_boxplot(aes(fill = arm), outlier.shape = NA, alpha = 0.4, width = 0.4) +
  geom_point(aes(color = arm), size = 2.5, alpha = 0.7) +
  facet_wrap(~ arm) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  scale_color_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = iners_pvals,
    aes(x = x_pos, y = max_val, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Lactobacillus iners",
    subtitle = "Paired change from baseline to after antibiotics",
    x        = NULL,
    y        = "Bacterial load (copies/mL)"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "none"
  )
```

---

## Master summary: All bacteria combined

```{r master-table}
# Build one big table for all bacteria
build_bacteria_table <- function(bacteria_name) {
  wide <- df_long %>%
    filter(bacteria == bacteria_name) %>%
    select(pid, arm, time_point, value, detected) %>%
    pivot_wider(
      names_from  = time_point,
      values_from = c(value, detected),
      names_glue  = "{time_point}_{.value}"
    )

  detect <- wide %>%
    group_by(arm) %>%
    summarise(
      n_total            = n(),
      baseline_detected  = sum(Baseline_detected == "Detected"),
      baseline_pct       = round(baseline_detected / n_total * 100, 0),
      after_detected     = sum(`After antibiotic_detected` == "Detected"),
      after_pct          = round(after_detected / n_total * 100, 0),
      .groups = "drop"
    ) %>%
    mutate(
      baseline_lbl = paste0(baseline_detected, "/", n_total, " (", baseline_pct, "%)"),
      after_lbl    = paste0(after_detected, "/", n_total, " (", after_pct, "%)")
    )

  summary <- wide %>%
    group_by(arm) %>%
    summarise(
      bl_med_iqr = paste0(
        format(median(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
        " (",
        format(IQR(Baseline_value, na.rm = TRUE), big.mark = ",", scientific = FALSE),
        ")"
      ),
      aft_med_iqr = paste0(
        format(median(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
        " (",
        format(IQR(`After antibiotic_value`, na.rm = TRUE), big.mark = ",", scientific = FALSE),
        ")"
      ),
      .groups = "drop"
    )

  pval <- wide %>%
    group_by(arm) %>%
    summarise(
      p = wilcox.test(Baseline_value, `After antibiotic_value`, paired = TRUE, exact = FALSE)$p.value,
      .groups = "drop"
    ) %>%
    mutate(p_lbl = formatC(p, format = "e", digits = 2))

  detect %>%
    left_join(summary, by = "arm") %>%
    left_join(pval, by = "arm") %>%
    mutate(bacteria = bacteria_name, .before = 1)
}

master <- bind_rows(
  build_bacteria_table("Total bacteria"),
  build_bacteria_table("L. crispatus"),
  build_bacteria_table("L. iners")
) %>%
  select(
    Bacteria = bacteria,
    Arm = arm,
    N = n_total,
    "Baseline detected, n (%)" = baseline_lbl,
    "Baseline median (IQR)" = bl_med_iqr,
    "After antibiotic detected, n (%)" = after_lbl,
    "After antibiotic median (IQR)" = aft_med_iqr,
    "Paired p-value" = p_lbl
  )

knitr::kable(master, caption = "Table 4: Complete summary — all bacteria, detection rates, levels, and paired p-values")
```

## Bonus: All three bacteria in one figure

```{r plot-all-bacteria}
#| fig-width: 10
#| fig-height: 8

# Compute all paired p-values at once using the wider format
all_paired <- df_long %>%
  select(pid, arm, bacteria, time_point, value) %>%
  pivot_wider(names_from = time_point, values_from = value)

all_pvals <- all_paired %>%
  group_by(bacteria, arm) %>%
  summarise(
    p = wilcox.test(Baseline, `After antibiotic`, paired = TRUE, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(label = paste0("p = ", formatC(p, format = "e", digits = 2)))

all_ymax <- df_long %>%
  group_by(bacteria, arm) %>%
  summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop")

all_pvals <- all_pvals %>% left_join(all_ymax, by = c("bacteria", "arm"))

ggplot(df_long, aes(x = time_point, y = value)) +
  geom_line(aes(group = pid), color = "grey60", alpha = 0.3, linewidth = 0.5) +
  geom_boxplot(aes(fill = arm), outlier.shape = NA, alpha = 0.4, width = 0.4) +
  geom_point(aes(color = arm), size = 1.8, alpha = 0.6) +
  facet_grid(bacteria ~ arm, scales = "free_y") +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  scale_color_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = all_pvals,
    aes(x = 2.3, y = max_val, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.2, color = "grey30"
  ) +
  labs(
    title    = "qPCR Bacteria — Paired Change Over Time",
    subtitle = "How each participant's bacteria changed from baseline to after antibiotic, by arm",
    x        = NULL,
    y        = "Bacterial load (copies/mL)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold", size = 11),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "bottom"
  )
```

---

## How to re-run this tomorrow

Open this file in RStudio and click the **Render** button (or
`Ctrl`/`Cmd` + `Shift` + `K`). Quarto will run all the code and create
an HTML file with the tables and plots.