---
title: "Group A Yogurt Study — Bacteria Box Plots"
subtitle: "Comparing unchanged diet vs yogurt arm at baseline and after antibiotics"
format: html
editor: source
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(ggpubr)
library(rstatix)
library(here)
```

## About this file

This Quarto document creates box plots for each bacteria measured in the
Group A Yogurt study. For each bacteria you'll see **two panels side by
side** — baseline on the left, after antibiotics on the right — each
comparing the **unchanged diet** arm against the **yogurt** arm with a
p-value.

### What the bacteria columns mean

| Column | What it measures |
|--------|-------------------|
| `qpcr_bacteria` | **Total bacteria** — all bacteria together |
| `qpcr_crispatus` | ***Lactobacillus crispatus*** — a "good" bacteria linked to vaginal health |
| `qpcr_iners` | ***Lactobacillus iners*** — another common species, less protective |

```{r load-data}
# Read the three data files
samples <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
meta    <- read_csv(here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
qpcr    <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))

# Preview the qPCR data to see what we're working with
glimpse(qpcr)
```

```{r merge-and-tidy}
# Merge qPCR values with time_point and arm info
df <- samples %>%
  left_join(qpcr, by = "sample_id")

# Clean up labels for nice-looking plots
df <- df %>%
  mutate(
    arm = recode(arm,
                 "unchanged_diet" = "Unchanged diet",
                 "yogurt"         = "Yogurt"),
    time_point = recode(time_point,
                        "baseline"         = "Baseline",
                        "after_antibiotic" = "After antibiotic"),
    # Put time points in the right order (baseline first)
    time_point = fct_relevel(time_point, "Baseline", "After antibiotic")
  )

# Pivot so each bacteria gets its own row
df_long <- df %>%
  pivot_longer(
    cols      = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to  = "bacteria",
    values_to = "value"
  ) %>%
  mutate(
    bacteria = recode(bacteria,
                      "qpcr_bacteria"  = "Total bacteria",
                      "qpcr_crispatus" = "L. crispatus",
                      "qpcr_iners"     = "L. iners"),
    # Keep the right order
    bacteria = fct_relevel(bacteria, "Total bacteria", "L. crispatus", "L. iners")
  )

head(df_long)
```

## Total bacteria

```{r plot-total-bacteria}
#| fig-width: 8
#| fig-height: 5

total_df <- df_long %>% filter(bacteria == "Total bacteria")

# Compute p-values for each time point (Wilcoxon test — good for small samples)
total_pvals <- total_df %>%
  group_by(time_point) %>%
  summarise(
    p = wilcox.test(value ~ arm, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Wilcoxon p = ", formatC(p, format = "e", digits = 2)),
    # Position the label at the top-right of each panel
    y_pos = total_df %>%
      group_by(time_point) %>%
      summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop") %>%
      pull(max_val)
  )

ggplot(total_df, aes(x = arm, y = value, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.5, color = "grey40") +
  facet_wrap(~ time_point) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  # Add p-value labels
  geom_text(
    data    = total_pvals,
    aes(x = 2.3, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Total bacteria",
    subtitle = "Unchanged diet vs Yogurt arm",
    x        = NULL,
    y        = "Bacterial load (copies/mL)",
    fill     = "Arm"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "bottom"
  )
```

**What this shows:** Total bacteria count in each arm, at baseline and
after antibiotics. Each dot is one participant. The box shows the
middle 50% of values, the line inside is the median.

---

## *Lactobacillus crispatus*

```{r plot-crispatus}
#| fig-width: 8
#| fig-height: 5

crispatus_df <- df_long %>% filter(bacteria == "L. crispatus")

# Compute p-values
crispatus_pvals <- crispatus_df %>%
  group_by(time_point) %>%
  summarise(
    p = wilcox.test(value ~ arm, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Wilcoxon p = ", formatC(p, format = "e", digits = 2)),
    y_pos = crispatus_df %>%
      group_by(time_point) %>%
      summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop") %>%
      pull(max_val)
  )

ggplot(crispatus_df, aes(x = arm, y = value, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.5, color = "grey40") +
  facet_wrap(~ time_point) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = crispatus_pvals,
    aes(x = 2.3, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Lactobacillus crispatus",
    subtitle = "Unchanged diet vs Yogurt arm",
    x        = NULL,
    y        = "Bacterial load (copies/mL)",
    fill     = "Arm"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "bottom"
  )
```

**What this shows:** *L. crispatus* levels. Notice many participants have
zero (0) — this bacteria is not present in everyone. The p-value tells
you if the difference between arms is bigger than what you'd expect by
chance.

---

## *Lactobacillus iners*

```{r plot-iners}
#| fig-width: 8
#| fig-height: 5

iners_df <- df_long %>% filter(bacteria == "L. iners")

# Compute p-values
iners_pvals <- iners_df %>%
  group_by(time_point) %>%
  summarise(
    p = wilcox.test(value ~ arm, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("Wilcoxon p = ", formatC(p, format = "e", digits = 2)),
    y_pos = iners_df %>%
      group_by(time_point) %>%
      summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop") %>%
      pull(max_val)
  )

ggplot(iners_df, aes(x = arm, y = value, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.5, color = "grey40") +
  facet_wrap(~ time_point) +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = iners_pvals,
    aes(x = 2.3, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.5, color = "grey30"
  ) +
  labs(
    title    = "Lactobacillus iners",
    subtitle = "Unchanged diet vs Yogurt arm",
    x        = NULL,
    y        = "Bacterial load (copies/mL)",
    fill     = "Arm"
  ) +
  theme_bw(base_size = 14) +
  theme(
    strip.text       = element_text(face = "bold", size = 13),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "bottom"
  )
```

---

## Bonus: All three bacteria in one figure

This puts all bacteria together so you can compare the patterns at a
glance.

```{r plot-all-bacteria}
#| fig-width: 10
#| fig-height: 8

# Compute all p-values at once
all_pvals <- df_long %>%
  group_by(bacteria, time_point) %>%
  summarise(
    p = wilcox.test(value ~ arm, exact = FALSE)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("p = ", formatC(p, format = "e", digits = 2)),
    y_pos = df_long %>%
      group_by(bacteria, time_point) %>%
      summarise(max_val = max(value, na.rm = TRUE) * 1.15, .groups = "drop") %>%
      pull(max_val)
  )

ggplot(df_long, aes(x = arm, y = value, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4, color = "grey40") +
  facet_grid(bacteria ~ time_point, scales = "free_y") +
  scale_fill_manual(values = c("Unchanged diet" = "#4A90D9", "Yogurt" = "#E8833A")) +
  geom_text(
    data    = all_pvals,
    aes(x = 2.3, y = y_pos, label = label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 3.2, color = "grey30"
  ) +
  labs(
    title    = "qPCR Bacteria — Yogurt vs Unchanged Diet",
    subtitle = "Comparison at baseline and after antibiotic",
    x        = NULL,
    y        = "Bacterial load (copies/mL)",
    fill     = "Arm"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold", size = 11),
    plot.title       = element_text(face = "bold", hjust = 0.5, size = 15),
    plot.subtitle    = element_text(hjust = 0.5, color = "grey40"),
    legend.position  = "bottom"
  )
```

---

## Quick summary table

```{r summary-table}
df_long %>%
  group_by(bacteria, time_point, arm) %>%
  summarise(
    n        = n(),
    mean     = mean(value, na.rm = TRUE) %>% round(0),
    median   = median(value, na.rm = TRUE) %>% round(0),
    min      = min(value, na.rm = TRUE) %>% round(0),
    max      = max(value, na.rm = TRUE) %>% round(0),
    .groups  = "drop"
  ) %>%
  mutate(across(where(is.numeric), ~ format(., big.mark = ",")))
```

---

## How to re-run this tomorrow

Open this file in RStudio and click the **Render** button (or
`Ctrl`/`Cmd` + `Shift` + `K`). Quarto will run all the code and create
an HTML file with the plots. You can also knit individual chunks by
clicking the green play arrow on each code block.
