---
title: "Twin Paired Box Plots — Male & Female Twins Over Time"
author: "Intro to R for Data Science with LLMs"
format: html
editor: source
---

## Overview

This notebook simulates **twin paired data**: male–female twin pairs measured at
baseline (visit 0) and up to 10 follow-up visits. The discrepancy between male
and female twins varies across visits — some visits show little difference, 
others show clearer separation.

The data is split into **long format** so we can easily facet by visit and
compare the distributions of male vs female twins at each time point.

```{r setup, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggpubr)
library(rstatix)

set.seed(42)  # reproducible

knitr::opts_chunk$set(fig.width = 12, fig.height = 7, dpi = 200)
```

---

## 1. Simulate twin paired data

Each **twin pair** shares a genetic baseline. Then we add:
- A **sex effect** (male or female)
- A **visit effect** that changes over time
- A **time × sex interaction** so the male–female gap varies per visit
- Random noise

```{r simulate}
n_pairs   <- 25           # number of twin pairs
n_visits  <- 11           # visits 0 through 10
total_obs <- n_pairs * 2 * n_visits  # 2 sexes × 11 visits

# ── Per-pair genetic baseline (shared by both twins) ─────
pair_id <- rep(1:n_pairs, each = 2 * n_visits)
genetic_baseline <- rep(rnorm(n_pairs, mean = 50, sd = 10),
                        each = 2 * n_visits)

# ── Sex label (each pair: one male, one female) ──────────
# For each pair, first half of rows = female, second half = male
sex <- rep(rep(c("Female", "Male"), each = n_visits), times = n_pairs)

# ── Visit number (0 to 10) ────────────────────────────────
visit <- rep(0:(n_visits - 1), times = n_pairs * 2)

# ── Sex effect: males start slightly higher ──────────────
sex_effect <- ifelse(sex == "Male", 5, 0)

# ── Visit effect: gentle upward trend ────────────────────
visit_effect <- visit * 0.5

# ── Time × sex interaction (varies per visit) ────────────
# We create a pattern where the male–female gap changes:
#   visit 0: small gap   visit 3: large gap   visit 6: reversed
#   visit 9: large gap   visit 10: small gap
interaction_pattern <- data.frame(
  visit = 0:10,
  interaction = c(0, 1, 3, 6, 4, 2, -2, 1, 4, 7, 2)
)
interaction_map <- interaction_pattern$interaction[match(visit, interaction_pattern$visit)]
time_x_sex <- ifelse(sex == "Male", interaction_map, 0)

# ── Random noise ──────────────────────────────────────────
noise <- rnorm(total_obs, mean = 0, sd = 4)

# ── Assemble the data frame ───────────────────────────────
twins <- data.frame(
  pair     = pair_id,
  sex      = sex,
  visit    = visit,
  value    = genetic_baseline + sex_effect + visit_effect + time_x_sex + noise
)

# Also create a wide format to show the pairing
twins_wide <- twins %>%
  pivot_wider(
    id_cols     = c(pair, visit),
    names_from  = sex,
    values_from = value,
    names_prefix = ""
  )

# Quick peek at the structure
glimpse(twins)
```

Let's look at a few twin pairs to see the pattern:

```{r peek, echo=FALSE}
twins_wide %>%
  filter(pair <= 4) %>%
  mutate(across(c(Female, Male), ~ round(., 1))) %>%
  knitr::kable(caption = "First 4 twin pairs across all visits (wide format)")
```

---

## 2. Box plot: male vs female at each visit

Side-by-side box plots for every visit, faceted so you can watch the gap
change over time.

```{r boxplot, warning=FALSE}
# Convert visit to a factor so ggplot treats it as a category
twins$visit <- factor(twins$visit, levels = 0:10,
                      labels = paste0("Visit ", 0:10))

p_box <- twins %>%
  ggplot(aes(x = sex, y = value, fill = sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.55, width = 0.5) +
  geom_jitter(aes(color = sex),
              width = 0.12, size = 1.2, alpha = 0.4) +
  facet_wrap(~ visit, nrow = 3) +
  scale_fill_manual(values = c("Female" = "#D95F69", "Male" = "#5B9BD5")) +
  scale_color_manual(values = c("Female" = "#D95F69", "Male" = "#5B9BD5")) +
  labs(
    title    = "Male vs Female Twins Over Time",
    subtitle = "Each panel = one visit. The male–female gap varies across visits.",
    x        = NULL,
    y        = "Measurement value",
    fill     = "Sex",
    color    = "Sex"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    legend.position = "bottom"
  )

print(p_box)
```

---

## 3. Paired box plot — connecting each twin pair

This version draws **paired points** connected by lines so you can see the
within-pair difference (dots connected with a line).

```{r paired-boxplot, warning=FALSE}
# Only show a subset of visits to avoid overcrowding
visits_to_show <- c("Visit 0", "Visit 2", "Visit 4", "Visit 6", "Visit 8", "Visit 10")

# Use the long-format data directly.
# geom_line(aes(group = pair)) connects the two points (Female, Male) within each twin pair.
p_paired <- twins %>%
  filter(visit %in% visits_to_show) %>%
  ggplot(aes(x = sex, y = value)) +
  geom_boxplot(aes(fill = sex), outlier.shape = NA, alpha = 0.4, width = 0.4) +
  geom_line(aes(group = pair), color = "grey50", alpha = 0.5, linewidth = 0.3) +
  geom_point(aes(color = sex), size = 1.5, alpha = 0.7) +
  facet_wrap(~ visit, nrow = 2) +
  scale_fill_manual(values = c(Female = "#D95F69", Male = "#5B9BD5")) +
  scale_color_manual(values = c(Female = "#D95F69", Male = "#5B9BD5")) +
  labs(
    title    = "Paired Twin Data — Each Line Connects One Twin Pair",
    subtitle = "Male–female differences within each twin pair. Gap changes over visits.",
    x        = NULL,
    y        = "Measurement value",
    fill     = "Sex",
    color    = "Sex"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    legend.position = "bottom"
  )

print(p_paired)
```

---

## 4. Statistical tests per visit

Let's check how the p-values (t-test and Wilcoxon) change across visits,
reflecting the varying discrepancy between male and female twins.

```{r pvalues, warning=FALSE}
pvals <- twins %>%
  group_by(visit) %>%
  summarise(
    n_female    = sum(sex == "Female"),
    n_male      = sum(sex == "Male"),
    mean_female = mean(value[sex == "Female"], na.rm = TRUE),
    mean_male   = mean(value[sex == "Male"], na.rm = TRUE),
    diff        = mean_male - mean_female,
    t_pval      = tryCatch(t.test(value ~ sex)$p.value,
                           error = function(e) NA),
    wilcox_pval = tryCatch(wilcox.test(value ~ sex)$p.value,
                           error = function(e) NA),
    .groups = "drop"
  ) %>%
  mutate(
    across(ends_with("pval"), ~ round(., 4)),
    t_label      = paste0("t-test p = ", t_pval),
    wilcox_label = paste0("Wilcoxon p = ", wilcox_pval)
  )

knitr::kable(pvals %>% select(visit, n_female, n_male, diff, t_label, wilcox_label),
             caption = "Male–female difference and p-values at each visit")
```

---

## 5. Summary plot with p-values on the panels

Finally, a cleaner box plot with p-values printed directly on each panel.

```{r summary-plot, warning=FALSE}
# Prepare annotation data
pvals_annot <- pvals %>%
  mutate(
    combined_label = paste0(t_label, "\n", wilcox_label),
    # Position text in the top-right of each panel
    x_pos = 2.3
  )

# Find y_max for each panel
y_max_vals <- twins %>%
  group_by(visit) %>%
  summarise(y_max = max(value, na.rm = TRUE) * 1.15, .groups = "drop")

pvals_annot <- pvals_annot %>%
  left_join(y_max_vals, by = "visit")

p_summary <- twins %>%
  ggplot(aes(x = sex, y = value, fill = sex)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.55, width = 0.5) +
  geom_jitter(width = 0.12, size = 1.2, alpha = 0.4, color = "grey30") +
  geom_text(
    data = pvals_annot,
    aes(x = 2.3, y = y_max, label = combined_label),
    inherit.aes = FALSE,
    hjust = 1, vjust = 1, size = 2.8, color = "grey40"
  ) +
  facet_wrap(~ visit, nrow = 3) +
  scale_fill_manual(values = c("Female" = "#D95F69", "Male" = "#5B9BD5")) +
  labs(
    title    = "Male vs Female Twins — With Statistical Tests",
    subtitle = "t-test and Wilcoxon p-values shown per visit",
    x        = NULL,
    y        = "Measurement value",
    fill     = "Sex"
  ) +
  theme_bw(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "grey40"),
    legend.position = "bottom"
  )

print(p_summary)
```

---

**Key takeaways:**

- The **gap between male and female twins changes over visits** — some visits
  show a big difference (e.g., visits 3, 9, 10), others show almost none
  (e.g., visit 6 where the gap reverses direction).
- The **paired lines** in section 3 highlight that these are *matched* pairs
  sharing a genetic baseline — so the within-pair difference is the cleanest
  signal.
- The p-values in section 4 confirm: at some visits the difference is
  statistically significant, at others it isn't, matching the varying
  discrepancy pattern we built into the simulation.